For Finitary Induction-Induction, \\ Induction is Enough

### 2020-04-23 paper/paper.tex
some intro

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}

@@

\title{For finitary induction-Induction, Induction is Enough} %TODO Please add

@@

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the James
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by T. Altenkirch, A. Kaposi, and A. Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.
\end{abstract}

@@

Many mutual inductive types can be reduced to indexed inductive types, where the
index disambiguates different sorts. An example is the mutual inductive datatype
with two sorts
$\mathsf{isEven}$ and $\mathsf{isOdd}$, defined by the following
constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(n+1) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(n+1)
\end{alignat*}
These can be reduced to the following single inductive family.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(n+1) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(n+1)
\end{alignat*}

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimal use case
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also shows up in higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions, where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs\footnote{An experimental version of Coq with IITs is also
  available on github.}.

We now state  
the elimination principle for $\Con$--$\Ty$.
We implicitly universally quantify over the variables $\Gamma$ of type $\Con$
and $A$ of type $\Ty$ when required (later in the development, we may similarly omit universal quantification).
Given the following components,
\begin{alignat*}{5}
  & \Con^D && : \Con\ra\Set \\
  & \Ty^D && : \Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
  & \mathsf{ext}^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,A\ra \Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & \U^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & \El^D && : (\Gamma^D:\Con^D\,\Gamma)\ra\,\Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,(\U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:\Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra \Con^D\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra \Ty^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = \U^D\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = \El^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

Reducing IITs to inductive types is an open problem. Forsberg
\cite{forsberg-phd} showed a reduction which only supports a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper} reduced
several IITs to inductive types, working inside a cubical type theory, but he
also only constructed a weaker eliminator.

The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is a small
type theory in which every context encodes a valid quotient inductive-inductive
type (QIIT) signature. For example, the signature for natural numbers is simply
the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$
and $suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315} showed
that if this theory of signatures exists, then any finitary QIIT can be derived
from it via a term model construction.
Here, by finitary QIIT, we mean that infinitary constructors, such as the following
infinitary $\Pi$, are not allowed:
\[
  \Pi_\infty : (\Gamma : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\Gamma) \rightarrow \Ty\,\Gamma
   \]
If we omit the type former for equality
constructors from the theory of signatures, we get a theory of finitary IIT
signatures, and the construction of \cite{Kaposi:2019:CQI:3302515.3290315} still
works: constructing an arbitrary IIT from the theory of IIT signatures does not
need the $\mathsf{Id}$ type former to be present.

However, the QIIT term model construction assumes that we already have the
syntax of QIIT signatures, and we can do induction over it. Hence, it only shows
that all finitary QIITs are derivable from a particular fairly complicated QIIT.
Likewise, restricted to finitary IITs, we only know that they are constructible
from a particular QIIT.

In this paper, we construct the theory of IIT signatures, which is a large
infinitary QIIT, using only indexed inductive types, uniqueness of identity
types (UIP) and function extensionality. Thus, we show that all finitary IITs are
reducible to the same feature set.

@@

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{csquotes} %for \llbracket and \rrbracket
\usepackage{stmaryrd}

@@

\title{For Finitary Induction-Induction, \\ Induction is Enough} %TODO Please add

@@

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.
\end{abstract}

@@

Many mutual inductive types can be reduced to indexed inductive types, where the
index disambiguates different sorts. For example, consider the mutual inductive datatype
with two sorts
$\mathsf{isEven}$ and $\mathsf{isOdd}$, defined by the following
constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(\mathsf{suc}\,n) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(\mathsf{suc}\,n)
\end{alignat*}
This can be reduced to the following single inductive family where
$\mathsf{isEven?}\,\mathsf{true}$ represents $\mathsf{isEven}$ and
$\mathsf{isEven?}\,\mathsf{false}$ represent $\mathsf{isOdd}$.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(\mathsf{suc}\,n) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(\mathsf{suc}\,n)
\end{alignat*}

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimsed version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also shows up in higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,({ext}^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
Above we used implicit quantifications for $\Gamma:\Con$ and
$A:\Ty\,\Gamma$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\Gamma$ before its explicit parameter of type
$Con^D\,\Gamma$.

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = {ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and it is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
extensional type theory however this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\Gamma\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\Gamma\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^S && : Con^S\,\Gamma\ra Ty^S\,(\mathsf{U}\,\Gamma) \\
  & El^S && : Con^S\,\Gamma\ra\,Ty^S\,(\mathsf{El}\,\Gamma) \\
  & \mathsf{selimCon} && : (\Gamma:\Con)\ra Con^S\,\Gamma \\
  & \mathsf{selimTy} && : (A:\Ty\,\Gamma)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak to define e.g.\ the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\Gamma\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\Gamma\,A\rr && := (\gamma:\ll\Gamma\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\Gamma\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\Gamma\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle is that
$\ll\blank\rr$ for types refers to $\ll\blank\rr$ for contexts, hence
this function is recursive-recursive.

Kaposi, Kov{\'a}cs, and Altenkirch
\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory
called the theory of signatures to describe quotient
inductive-inductive types (QIIT). QIITs are generalisations of IITs
where equality constructors are also allowed. A QIIT signature is a
context in the theory of QIIT signatures, for example natural numbers
are specified by the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of
length three ($Nat$, $zero$ and $suc$ are variable names). The theory
of QIIT signatures is itself a QIIT. In ibid., it is proved that if a
model of extensional type theory supports the theory of QIIT
signatures, then it supports all QIITs.

By omitting the identity type former from the theory of QIIT
signatures, we obtain a theory of IIT signatures and the construction
is still valid. It follows that if a model of extensional type theory
supports the theory of IIT signatures, it supports all IITs. In this
paper we show that any model of extensional type theory with indexed
W-types supports the theory of signatures, hence all IITs.

Just as ibid., we only consider finitary IITs, that is, constructors
can only have a finite number of recursive arguments. An example
constructor for $\Con$--$\Ty$ which is not allowed is the following:
\[
  \Pi_\infty : (\Gamma : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\Gamma) \rightarrow \Ty\,\Gamma
\]

TODO: list of contributions here.

@@

 (later in the development, we may similarly omit universal quantification)
